Constraining the Interiors of Asteroids Through Close Encounters

### 2022-04-14 paper/main.tex
Ready to stage new sigma scan

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{colortbl}
\usepackage{dsfont}
\usepackage{bm}
\usetikzlibrary{arrows.meta}

@@

% Abstract of the paper
\begin{abstract}
  We investigate the degree to which asteroid interior density distributions can be extracted from rotational velocity data gathered during a close encounter. We derive the equations of motion for an rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic rotational velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we re-extract the density moments of the asteroid in a wide range of scenarios to measure the degree to which best-fitting parameter precision is affected. Specifically, we use many injection-retrieval tests to study fit precision's dependence on the asteroid's moment of inertia, observational precision and cadence, orbital parameters, and initial spin pole direction as well as the quantity of near-pericentre data and the central body oblateness. Finally, we discuss the degeneracy between the density moments and the actual density distribution and propose three models to construct a representative density distribution from fit results.
\end{abstract}

@@

Over the past twenty years, the increase in quantity and quality of high-sensitivity, all-sky surveys has prompted the discovery of numerous asteroids. Such advances have been made via ground-based surveys such as the Catalina Sky Survey \cite{larson1998catalina}, Pan-STARRS \cite{kaiser2002pan}, and the Lincoln Near-Earth Asteroid Research project (LINEAR) \cite{stokes2000lincoln}, as well as space-based instruments such as the Wide-field Infrared Survey Explorer (WISE) mission \cite{wright2010wide}. Many of these asteroids are relatively small, but some are kilometre-sized and a few are predicted to encounter Earth or other planets closely in the near future. Light curve analysis has been performed on many of these new asteroids to study their surface properties, but analyses that are sensitive to the interior structure of the asteroid are rarer.

Theoretical work has demonstrated that tidal torque alters the angular velocity of an asteroid during a gravitational encounter with another body, and that this tidal torque is sensitive to the interior density distribution of the asteroid \cite{paul88, SCHEERES2000106, ashenberg07, BOUE2009750, HouMar2017}. This two-body generalized system has been studied to different orders and with several different methods. The angular velocity changes induced by tidal torque occur both in asteroid binaries \cite{Naidu_2015, Makarov2022ChaosOO}, and during planetary encounters \cite{RICHARDSON199847, scheeres2004evolution}. Such asteroids need not be tidally disrupted for their rotation state to be perturbed by tidal torque. Some examples of rotational velocity altered during an encounter have recently been observed, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and have been predicted for the 2029 encounter of (99942) Apophis with Earth \cite{SCHEERES2005281, DEMARTINI201993}. The upcoming encounter of Apophis is of particular interest and much attention has been devoted to accurate estimates of its orbital and physical properties \cite{yu2014numerical, hirabayashi2021finite,valvano2022apophis, Lee2022Apophis}, and in particular how its angular velocity will be affected by tidal torque \cite{souchay2014rotational, souchay2018changes}. Altered rotational states altered by Solar tidal torque have been excluded for 1I/2017 U1 `Oumuamua \cite{KWIECINSKI2018170}.

Such asteroid encounters have further been shown to reveal the physical properties of the encountering asteroid. Refs.~\cite{MOSKOVITZ2020113519, benson2020spin} extract best-fitting dimensions for a triaxial ellipsoid model of Duende from light-curve data, and Refs.~\cite{DESCAMPS2020113726} and \cite{BERTHIER2020113990} extract bulk densities and rotational periods for asteroid binaries (3905) Doppler and (617) Patroclus respectively.

More encounter candidates are likely to be discovered in the future --- for Earth or other planets --- potentially by the James Webb Space Telescope (JWST) or Large-aperture Synoptic Survey Telescope (LSST), and general methods for assessing whether an asteroid's encounter properties are conducive to the extraction of physical asteroid characteristics from light-curve data is necessary. However, such general encounters are not very well studied, since much of previous research has focused on specific encounters.

Furthermore, asteroid properties which appear in the beyond-leading-order terms of tidal torque have not yet been extracted, and more research is needed to study in what cases these effects are observable.

In this paper, we address these two topics by deriving a new equation of motion for an asteroid encounter system in section \ref{sec:methods}. This equation is simple and exact to arbitrary order, given assumptions reasonable for a planetary encounter (and many binary asteroid systems). It is also fast in that it allows volume integrals over the asteroid to be pre-computed. We use this equation of motion to simulate a generic asteroid encounter of Earth, and extract physical characteristics of the asteroid (specifically, density moments) from synthetic data via a Markov Chain Monte Carlo (MCMC) fit, described in section \ref{sec:fit}. We measure the uncertainties in the posterior distribution of density moments as a function of various encounter parameters such as the observational uncertainty, cadence of observation, encounter orbital parameters, and more. We express which parameters most affect the precision of fit results in section \ref{sec:results}. Finally, we propose methods to extract full density distributions with uncertainties from these density moments in section \ref{sec:distros}. We test the behaviour of these methods, highlight their strengths and weaknesses, and suggest ways in which they may be altered to suit the needs of the user.

@@

In this section, we describe our coordinates (section \ref{sec:coordinates}) for an encountering asteroid's position and orientation, and we parametrize its density distribution via its ``density moments'' (section \ref{sec:moments}). Then we derive an arbitrary-order equation for tidal torque (section \ref{sec:tidal-torque}), write the equations of motion for the system (section \ref{sec:eom}), and describe a simulation used to integrate the equation of motion and produce synthetic data of angular velocity over time (section \ref{sec:sim}).

\subsection{Coordinates}
\label{sec:coordinates}

Throughout this paper, we assume that the asteroid under study is on a hyperbolic encounter with periapse $r_p$ and excess velocity $v_\infty$. We do not consider any third-body perturbations, and we assume that the body being encountered (the central body, e.g.~a planet) is much more massive than the asteroid.

We make use of two frames of reference to model this system. One is the ``inertial frame,'' with axes denoted by $\unit{X}$, $\unit{Y}$, $\unit{Z}$ and origin placed at the central body's centre of mass and each axis aligned with an asteroid principal axis. $\unit{X}$ points from the central body to the asteroid periapse and $\unit{Z}$ points parallel to the orbit angular momentum. We will assume that the mass distribution of the central body is known in this inertial frame.

Our second frame is the ``body-fixed'' frame, denoted by $\unit{x}, \unit{y}, \unit{z}$. This frame is fixed with respect to the asteroid's principal axes and rotates with the asteroid, with its origin at the asteroid's centre of mass. We will solve for the asteroid's mass distribution with reference to the body-fixed frame. For definiteness, we define $\unit{z}$ to be the principal axis with maximal MOI. In general, we use capital letters to denote vectors in the inertial frame and lowercase vectors to denote vectors in the body-fixed frame.

The difference between the origins of the body-fixed and inertial frames is the position of the asteroid. We represent the relative orientations by $z-y-z$ Euler angles $\alpha$, $\beta$, and $\gamma$, such that a matrix $M$ rotating from the body-fixed to the inertial frame ($M\bm{r} = \bm{R}$) is given by
\begin{equation}
M = R_z(\alpha) R_y(\beta) R_z(\gamma).
\label{eqn:euler-angles}
\end{equation}
Here, $R_i(\theta)$ is a rotation around the unit vector $i$ by $\theta$ (figure \ref{fig:euler-angles}).

\begin{figure}
    \centering
    \begin{tikzpicture}
    \draw[-{Latex[length=3mm]}] (0, 0) -- (-4, 0) node[anchor=east] {$\unit x$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (2, -3) node[anchor=west] {$\unit y$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (0, 4) node[anchor=south] {$\unit z$};
    \draw[dashed, -{Latex[length=3mm]}] (0, 0) -- (3.7, -2) node[anchor=north] {$\unit n$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-0.5, -3) node[anchor=north] {$\unit X$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (4, 1) node[anchor=south] {$\unit Y$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-1.6, 2.7) node[anchor=south] {$\unit Z$};
    \draw[->] (0.5, -0.75) arc (290:302:2.5);
    \draw (0.9, -0.9) node[anchor=center] {$\alpha$};
    \draw[->] (0, 1.2) arc (130:161:1.3);
    \draw (-0.3, 1.3) node[anchor=center] {$\beta$};
    \draw[->] (0.97, -0.52) arc (330:368:1.3);
    \draw (1.4, -0.25) node[anchor=center] {$\gamma$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (0, 4) node[anchor=south] {$\unit Z$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (4, 0) node[anchor=west] {$\unit Y$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (-2, -2) node[anchor=east] {$\unit X$};
    \end{tikzpicture}
    \caption{$z-y-z$ Euler angles used in this work to express the orientation of the asteroid. Orientation is expressed as a rotation from the body-fixed axes (lowercase) to the inertial axes (bold and uppercase). The origins are co-located for demonstration purposes.}
    \label{fig:euler-angles}
\end{figure}

\subsection{Parameters: density moments}
\label{sec:moments}

In the next section, it will be shown that only certain parameters of the asteroid density distribution affect tidal torque called ``density moments.'' First, we define the un-normalized spherical harmonics $Y_{\ell m}(\theta, \phi) = P_{\ell m}(\cos \theta)e^{im\phi}$, where $P_{\ell m}$ are the associated Legendre Polynomials without the Condon-Shortley phase. The regular and irregular spherical harmonics then defined as:
\begin{equation}
  \begin{split}
    S_{\ell m}(\bm r) &= (-1)^m (\ell - m)! \frac{Y_{\ell m}(\unit r)}{r^{\ell+1}} \\
    R_{\ell m} (\bm r) &= (-1)^m \frac{r^\ell}{(\ell + m)!} Y_{\ell m}(\unit r).
  \end{split}
\end{equation}
These spherical harmonics obey many useful identities summarized in Ref.~\cite{Gelderen1998TheSO}, which derives spherical harmonic identities useful for quantum mechanics.

We define the density moments of an asteroid as
\begin{equation}
  K_{\ell m} = \frac{1}{\mu_\mathcal{A} a_\mathcal{A}^\ell} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r).
  \label{eqn:klm}
\end{equation}
Note that these are complex. Here, $\mathcal{A}$ indicates the volume of the asteroid, $\mu_\mathcal{A}$ is the mass of the asteroid, $\rho_\mathcal{A}(\bm r)$ is the density distribution, and $a_\mathcal{A}$ is a real length scale defined (for reasons covered later) as 
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{\mu_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2.
  \label{eqn:am}
\end{equation}
This length scale can be thought of as akin to the radius of the asteroid (although a spherical, uniform-density asteroid has radius $a_\mathcal{A}\sqrt{5/3} $). Both equations \ref{eqn:klm} and \ref{eqn:am} should be computed in the body-fixed axes.

Equations \ref{eqn:klm} and \ref{eqn:am} can be extended to the central body:
\begin{equation}
  \begin{split}
    &J_{\ell m} = \frac{1}{\mu_\mathcal{B} a_\mathcal{B}^\ell} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) R_{\ell m}(\bm r)\\
    &a_\mathcal{B}^2 = \frac{1}{\mu_\mathcal{B}} \int_\mathcal{B} d^3 r \rho_\mathcal{B}(\bm r) r^2.
  \end{split}
  \label{eqn:jlm}
\end{equation}
which should be computed in the inertial axes.

Note that both $J_{\ell m}$ and $K_{\ell m}$ are unitless. We call them ``moments'' because the $R_{\ell m}(\bm r)$ contains an $r^\ell$ dependence so that $K_{\ell m}$ is the $\ell$th density moment of the asteroid. The gravitational potential field of the asteroid can be written entirely in terms of $K_{\ell m}$ and $a_\mathcal{A}$, so we expect not to need any information about the density distribution of the asteroid beyond these parameters to compute tidal torque.

These moments share several key properties which we discuss before continuing. Firstly, for real mass density, properties of the spherical harmonics imply that $K_{\ell m} = (-1)^m K_{\ell, -m}^*$. Therefore, the set of $K_{\ell m}$ for $\ell < \ell_\text{max}$ contains $\ell_\text{max}^2$ degrees of freedom. However, some of these degrees of freedom are redundant with the choice of coordinates as we discuss next.

By definition, $K_{00}=1$. Furthermore, $K_{1m} = 0$ since the body-fixed frame is centred on the asteroid centre of mass. Further calculation reveals that the alignment of the body-fixed frame with the asteroid principal axes also forces $K_{21}= 0$ and $\Im K_{22}=0$ and the same for $m<0$. The only physical density moments for $\ell \leq 2$ are therefore $K_{22}$ and $K_{20}$, which are related to the moment of inertia around each principal axis by
\begin{equation}
  \begin{split}
    I_x &= \frac{2}{3}\mu_\mathcal{A} a_\mathcal{A}^2 \parens{K_{20} - 6 K_{22} + 1}\\
    I_y &= \frac{2}{3}\mu_\mathcal{A} a_\mathcal{A}^2 \parens{K_{20} + 6 K_{22} + 1}\\
    I_z &= \frac{2}{3}\mu_\mathcal{A} a_\mathcal{A}^2 \parens{-2K_{20} + 1}.
  \end{split}
  \label{eqn:moi}
\end{equation}
Incidentally, the definition of $a_\mathcal{A}$ was chosen to satisfy equation \ref{eqn:moi}.

The physical meaning of $K_{22}$ and $K_{20}$ can also be interpreted via a special case: if the asteroid is a uniform-density triaxial ellipsoid, the moments of inertia are simple to compute in terms of the semi-axis lengths and can be compared to those found in equation \ref{eqn:moi}. This yields semi-axis lengths of 
\begin{equation}
  \begin{split}
  a &= \sqrt{\frac{5}{3}}a_\mathcal{A}\sqrt{1-2K_{20}+12K_{22}}\\
  b &= \sqrt{\frac{5}{3}}a_\mathcal{A}\sqrt{1-2K_{20}-12K_{22}}\\
  c &= \sqrt{\frac{5}{3}}a_\mathcal{A}\sqrt{1+4K_{20}}.
  \label{eqn:ellipsoid-axes}
  \end{split}
\end{equation}

The physical meaning of the higher-order moments $K_{3m}$ can be aided by assessing their symmetry properties. An asteroid that is mirror-symmetric along the $\unit{x}$ axis (meaning $\rho_\mathcal{A}(x,y,z)=\rho_\mathcal{A}(-x,y,z)$) necessarily sets certain density moments to zero. Which density moments are zeroed by which mirror symmetries is outlined in table \ref{tab:klm-symmetries}. Note that, while no mirror symmetries set $K_{00}$, $K_{20}$, or $K_{22}$ equal to zero, mirror symmetries exist which zero all the other moments, including $K_{3m}$. $\Re K_{32}$, $K_{31}$, and $K_{30}$ are the only $K_{3m}$ components zeroed by only one axis. This will not affect our fit results for $K_{3m}$, but when we compute density moments for sample distributions, we find that most $K_{3m}=0$.

\begin{table}
  \centering
  \begin{tabular}{c|ccccccc}
    \hline
    $\ell$ & $\Re K_{\ell 3}$ & $\Im K_{\ell 3}$ & $\Re K_{\ell 2}$ & $\Im K_{\ell 2}$ & $\Re K_{\ell 1}$ & $\Im K_{\ell 1}$ & $K_{\ell 0}$ \\ \hline
    0 &  &  &  &  &  &  & -\\ 
    1 &  &  &  &  & x & y & z\\ 
    2 &  &  & - & x,y & y,z & x,z & -\\ 
    3 & x,z & y,z & z & x,y,z & x & y & z\\ \hline
  \end{tabular}
  \caption{Axes of mirror symmetry that imply zeroed density moments. For example, for mirror symmetries along $\unit y$ or $\unit z$, $\Im K_{32}=0$. Mirror symmetry along $\unit x$ means $\rho_\mathcal{A}(x, y, z) = \rho_\mathcal{A}(-x, y, z)$. Dashes indicate that none of the mirror symmetries zero the moment in question. Since $r^2>0$ for $r\neq 0$, no symmetries set $a_\mathcal{A}=0$ either.}
  \label{tab:klm-symmetries}
\end{table}

Finally, the requirement that $\rho_\mathcal{A}(\bm r) \geq 0$ everywhere restricts $K_{\ell m}$. In the case of $K_{2m}$, this fact and the constraint that $I_z$ is larger than $I_x$ or $I_y$ requires $K_{20}$ and $K_{22}$ to fall in the triangle
\begin{equation}
  -\frac{1}{4} \leq K_{20} \leq 0, \qquad |K_{22}| \leq -\frac{K_{20}}{2}.
  \label{eqn:parameter-bounds}
\end{equation}
In practice, we also observe that $|K_{3m}| < 1$; often, $|K_{3m}| < 0.01$ even. For the rest of this paper, we therefore treat an uncertainty on $|K_{3m}|$ of $0.01$ or less as a practical limit on observability, in that uncertainty greater than 0.01 means $K_{3m}$ are effectively unknown. This is only a rough estimate, however.

\subsection{Tidal torque}
\label{sec:tidal-torque}

Derivations for the tidal torque experienced by a rigid body in the gravitational field of a larger mass have been computed by several previous studies \cite{paul88,HouMar2017,BOUE2009750, ashenberg07}, often in terms of the moment of inertia of the rigid body (or higher order moments of inertia), and to varying degrees of precision. A simple, first-order derivation is also easily computable in terms of the asteroid moment of inertia in the inertial frame.

Here, we present a novel derivation of the tidal torque to arbitrary orders in terms of the density moments of an asteroid defined in section \ref{sec:moments}. These density moments can be pre-computed and do not have to be re-evaluated every time-step.

Throughout this paper, we assume that the asteroid remains rigid throughout the encounter. We also assume no third-body perturbations from other Solar System objects. (Actually, third-body perturbing objects are allowed if they are closer to the central body's centre of mass than the asteroid perigee distance. Then, their density moments can be included in the density moments of the central body and this derivation can still be used.) For the sake of simplicity, we also assume that the density moments of the central body are known and do not evolve with time (i.e., the central body does not rotate).

The gravitational potential energy of the central body is, in its most general form,
\begin{equation}
V(\bm R') = -G\int_\mathcal{B} d^3 R \rho_\mathcal{B}(\bm R) \frac{1}{|\bm{R}-\bm{R'}|}.
\label{eqn:first-pe}
\end{equation}
where $\rho_\mathcal{B}$ is the density distribution of the central body and $\mathcal{B}$ indicates the central body's volume. All vectors here are written in the inertial frame. Given $|\bm{R}| < |\bm{R'}|$, Ref.~\cite{Gelderen1998TheSO} gives the identity
\begin{equation}
  \frac{1}{|\bm R - \bm R'|} = \sum_{\ell, m} R_{\ell m}(\bm R) S_{\ell m}^*(\bm R'),
  \label{eqn:ylm-expansion}
\end{equation}
where the sum is shorthand for $\sum_{\ell, m} = \sum_{\ell = 0}^\infty \sum_{m=-\ell}^\ell$.
We are interested in translating the potential energy of equation \ref{eqn:first-pe} to the body-fixed frame. To do this, we let $\bm{R'} = \bm D + \bm U$, where $\bm D$ is the location of the asteroid in the inertial frame. We further define $\bm U = M\bm u$, where $\bm u$ is in the body-fixed frame and $M$ is the rotation matrix given by the Euler angles $\alpha$, $\beta$, and $\gamma$ (see section \ref{sec:coordinates}). The translation from $\bm {R'}$ to $\bm U$ is then attained by the identity 
\begin{equation}
  S_{\ell m}(\bm R') = \sum_{\ell', m'} (-1)^{\ell'}R^*_{\ell' m'}(\bm U)S_{\ell+\ell', m + m'} (\bm D),
  \label{eqn:ylm-translation}
\end{equation}  
provided by Ref.~\cite{Gelderen1998TheSO}, and from $\bm U$ to $\bm u$ is given by
\begin{equation}
  \begin{split}
    Y_{\ell m}(M\bm u) = \sum_{m'=-\ell}^\ell & (-1)^{m+m'}\sqrt{\frac{(\ell-m')!(\ell+m)!}{(\ell+m')!(\ell-m)!}} \\
    & \times \mathcal{D}^\ell_{mm'}(M)^* Y_{\ell m'}(\bm u).\\
  \end{split}
  \label{eqn:ylm-rotation}
\end{equation}
Here, $\mathcal{D}^\ell_{mm'}(M)$ are the Wigner-$D$ matrices, which are determined by the Euler angles $\alpha$, $\beta$, and $\gamma$ of $M$.

Equations \ref{eqn:first-pe} to \ref{eqn:ylm-rotation} then provide formula for $V(\bm u)$ expressed as a sum of integrals over $\mathcal{B}$ of the central body density $\rho_\mathcal{B}(\bm R)$ times $R_{\ell m}(\bm R)$. These are expressed via equation \ref{eqn:jlm} as $J_{\ell m}$.

The tidal torque experienced by the asteroid (in the body-fixed frame) is given by
\begin{equation}
  \bm{\tau}(\bm u) = \int_\mathcal{A} d^3 u \rho_\mathcal{A}(\bm u) (\bm u \times (-\nabla_{\bm u} V(\bm u)))
\end{equation}
where $\rho_\mathcal{A}$ is the density distribution of the asteroid and $\mathcal{A}$ indicates the volume of the asteroid. Making use of one more identity concerning the derivatives of spherical harmonics:
\begin{equation}
  \begin{split}
  \bm u \times \nabla R_{\ell m}(\bm u)=&\frac{1}{2}\Big[(i\unit x - \unit y)(\ell-m+1)R_{\ell,m-1}(\bm u)\\
  &+(i\unit x+\unit y)(\ell+m+1)R_{\ell,m+1}(\bm u)\\
  & +2im\unit z R_{\ell m}(\bm u)\Big],
  \end{split}
\end{equation}
tidal torque can now be expressed as a function only of $J_{\ell m}$, $K_{\ell m}$, $a_\mathcal{A}$, $a_\mathcal{B}$, and the asteroid orientation and position. Explicitly,
\begin{equation}
  \begin{split}
  \bm \tau = & G\frac{\mu_\mathcal{A}\mu_\mathcal{B}}{2}\left[\sum_{\ell, m} a_\mathcal{B}^\ell J_{\ell m} \sum_{\ell', m'}a_\mathcal{A}^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D) (-1)^{\ell'}\right.\\
  \times & \left.\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(\alpha, \beta, \gamma)^* \right. \\
  \times & \Big((i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} \\
  & +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\Big) \Bigg].
  \end{split}
  \label{eqn:tidal-torque}
\end{equation}
Some $K_{\ell m}$ terms are included in this equation with $|m|>\ell$; these should all be taken to be zero.

Equation \ref{eqn:tidal-torque} possesses a few explicit properties which we discuss before writing the asteroid equations of motion. Firstly, $K_{00}$ does not appear, so that $\bm \tau$ is independent of asteroid mass. The mean density of the asteroid is therefore not constrained by tidal torque analysis. Secondly, torque is largest when $D$ is small (as expected), with the leading order of $\bm \tau$ proportional to $D^{-3}$. Thirdly, each $J_{\ell m}K_{\ell' m'}$ term is multiplied by $(a_\mathcal{B}/D)^\ell (a_\mathcal{A}/D)^{\ell'}$, the latter of which especially is small in most cases. Equation \ref{eqn:tidal-torque} can therefore be computed approximately by removing terms of large $\ell$ and $\ell'$. For our analysis, we remove $\ell' > 3$ and we usually keep only $\ell=0$. Note that $\ell=1$ contributes nothing since $J_{1m}=0$.

Further insight can be gained by remarking the value of the first-order of $\bm \tau$ for particular Euler angle cases. Setting $\beta = 0$ produces diagonal Wigner-$D$ matrices, and hence $\bm \tau \parallel \unit z$ to first-order. Also, the component $\tau_z$ oscillates, so that for certain values of $\alpha$ and $\gamma$, $\bm \tau = 0.$ This $\beta=0$ condition is equivalent to $\unit z \parallel \unit Z$ (see figure \ref{fig:euler-angles}).

For $\beta = \pi/2$, there are two interesting cases. One is for $\alpha = \phi$ (or $\alpha = \pi + \phi$), where $\phi$ is the angle between the asteroid and the perigee. In this case, $\tau = 0$ to first-order. The second case is $\alpha = \phi \pm \pi/2$, when again $\bm \tau \parallel \unit z$ and $\tau_z$ oscillates. At perigee ($\phi=0$), these conditions are equivalent to $\unit z \parallel \unit X$ and $\unit z \parallel \unit Y$ respectively.

The $\bm \tau \parallel \unit z$ cases are interesting because they do not induce tumbling. If velocity is $\bm \omega \parallel \unit z$ (a non-tumbling state, since $\unit z$ is a principal axis), then $\bm \omega \parallel \bm L$ and $\bm \tau = \dot{\bm L} \parallel \dot{\bm \omega}$ so that $\bm \omega$ remains parallel to $\unit z$ and non-tumbling. These cases of torque are additionally significant because not as many terms contribute to $\tau_z$ as to $\tau_x$ and $\tau_y$.

\subsection{Equations of motion}
\label{sec:eom}

The equations of motion of the asteroid position $\bm D$ are given by Newton's law of gravitation:
\begin{equation}
  \dot{\bm V} = -\frac{G \mu_\mathcal{B}}{D^3} \bm D \qquad \dot{\bm D} = \bm V
  \label{eqn:pos-eom}
\end{equation}
Rather than derive equations of motion for the Euler angles (which suffer from gimbal lock), we instead represent the orientation of the asteroid with a quaternion $\quat q$ which can be converted into Euler angles to compute $\mathcal{D}(\alpha, \beta, \gamma)$. This quaternion evolves as 
\begin{equation}
  \dot{\quat q} = \frac{1}{2}\quat q\quat \omega.
  \label{eqn:quat-eom}
\end{equation}
for angular velocity $\bm \omega$ given in the body-fixed frame. The equations of motion of $\bm \omega$ in turn are given by
\begin{equation}
  \begin{split}
    I_x \dot \omega_1 - \omega_y \omega_z (I_y - I_z) &= \tau_x\\
    I_y \dot \omega_2 - \omega_z \omega_x (I_z - I_x) &= \tau_y\\
    I_z \dot \omega_3 - \omega_x \omega_y (I_x - I_y) &= \tau_z.
  \end{split}
  \label{eqn:omega-eom}
\end{equation}
Equations \ref{eqn:tidal-torque} to \ref{eqn:omega-eom} form a set of non-linear, first-order coupled differential equations in which can be numerically integrated. They are expressed in terms of the constant physical parameters $\mu_{M/m}$, $a_{M/m}$, $J_{\ell m}$ and $K_{\ell m}$ given the density moment-moment of inertia relations given by equation \ref{eqn:moi}.

Note that equation \ref{eqn:omega-eom} is independent of $a_\mathcal{A}$ to first-order in $\bm \tau$, because $I_{j} \propto a_\mathcal{A}^2$ for all $j$ and $\bm \tau \propto a_\mathcal{A}^2$. Therefore, scaling $a_\mathcal{A}$ merely scales the value of the sub-leading-order contributions to $\bm \tau$.

@@

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Reference asteroid configurations}
\label{app:reference-configs}

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{colortbl}
\usepackage{dsfont}
\usepackage{bm}

% For the sideways figure
\usepackage{pdflscape}
\usepackage{afterpage}

\usetikzlibrary{arrows.meta}

@@

% Abstract of the paper
\begin{abstract}
  Knowledge of the interior distribution of an asteroid can reveal its composition and constrain its evolutionary history. However, most observations are not sensitive to interior structure. We investigate the degree to which asteroid interior density distributions can be extracted from rotational velocity data gathered during a close encounter. We derive the equations of motion for an rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic rotational velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we re-extract the density moments of the asteroid in a wide range of scenarios to measure the degree to which best-fitting parameter precision is affected. Specifically, we use many injection-retrieval tests to study fit precision's dependence on the asteroid's moment of inertia, observational precision and cadence, orbital parameters, and initial spin pole direction as well as the quantity of near-pericentre data and the central body oblateness. Finally, we discuss the degeneracy between the density moments and the actual density distribution and propose three models to construct a representative density distribution from fit results.
\end{abstract}

@@

Over the past twenty years, the increase in quantity and quality of sensitive all-sky surveys has prompted the discovery of numerous asteroids. Such advances have been made via ground-based surveys such as the Catalina Sky Survey \cite{larson1998catalina}, Pan-STARRS \cite{kaiser2002pan}, and the Lincoln Near-Earth Asteroid Research project (LINEAR) \cite{stokes2000lincoln}, as well as space-based instruments such as the Wide-field Infrared Survey Explorer (WISE) mission \cite{wright2010wide}. Many of these asteroids are relatively small, but some are kilometre-sized and a few are predicted to encounter Earth or other planets closely in the near future. More encounter candidates are likely to be discovered in the future --- for Earth or other planets --- potentially by the Large-aperture Synoptic Survey Telescope (LSST) or global telescope networks such as the Las Cumbres Observatory (LCO). Light curve analysis has been performed on many of these new asteroids to study their surface properties, but observations that are sensitive to the interior structure of the asteroid are rarer.

However, tidal torque acting on an asteroid and perturbing its angular velocity during an encounter presents an opportunity to constrain the asteroid mass distribution. Such angular velocity perturbations have been observed and used to extract asteroid properties in several cases, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and asteroid binaries (3905) Doppler and (617) Patroclus respectively. Altered rotational states altered by Solar tidal torque have been excluded for 1I/2017 U1 `Oumuamua \cite{KWIECINSKI2018170}.

The upcoming encounter of Apophis is of particular interest and much theoretical work has been devoted to accurate estimates of its orbital and physical properties, and predictions of how its angular velocity will be affected by tidal torque \cite{yu2014numerical, hirabayashi2021finite,valvano2022apophis, Lee2022Apophis, souchay2014rotational, souchay2018changes}. This research builds on a larger body of theoretical work showing that the tidal torque, observed through angular velocity perturbations, is sensitive to asteroid interior density distribution \cite{Naidu_2015, Makarov2022ChaosOO, RICHARDSON199847, scheeres2004evolution}. This two-body generalized system has been studied to different orders and with several different methods \cite{paul88, SCHEERES2000106, ashenberg07, BOUE2009750, HouMar2017}. The success of this work is demonstrated by the recent extraction of moment of inertia ratios for Duende \cite{MOSKOVITZ2020113519}. However, density distribution information beyond the moment of inertia ratios have not yet been extracted for planetary encounters. More research is needed to study in what cases these effects are observable, and what factors generally inhibit observation of these sub-dominant terms.

In this paper, we address this question by deriving a new equation of motion for an asteroid encounter system, using , and describe a simulation used to integrate the equation of motion and produce synthetic data of angular velocity over time (section \ref{sec:sim}).it to simulate a generic asteroid encounter of Earth. In section  \ref{sec:methods}, we then extract physical characteristics of the asteroid (specifically, density moments) from synthetic data via a Markov Chain Monte Carlo (MCMC) fit. We also describe methods to extract estimates of density distribution from these density moments. In section \ref{sec:results}, we demonstrate our fit process and evaluate our density distribution extraction methods on density moments extracted from fits to synthetic data. In section \ref{sec:discussion}, we measure the uncertainties in the Posterior Probability Distribution (PPD) of density moments as a function of various encounter parameters such as the observational uncertainty, encounter orbital parameters, asteroid physical properties, and more.

@@

In appendix \ref{app:eom}, we derive a novel equation for the tidal torque experienced by an asteroid in a two-body system, accurate to arbitrary order. The equation is written in terms of the asteroid ``length'' 
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{\mu_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2
  \label{eqn:am}
\end{equation}
and (complex) ``density moments''
\begin{equation}
  K_{\ell m} = \frac{1}{\mu_\mathcal{A} a_\mathcal{A}^\ell} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r),
  \label{eqn:klm}
\end{equation}
both of which are constant and only need to be computed once to generate data. Here, $\rho_\mathcal{A}(\bm r)$ is the asteroid density distribution, $\mu_\mathcal{A}$ is the asteroid mass, and $R_{\ell m}$ are the regular solid spherical harmonics (see the appendix for details). They produce a tidal torque of 
\begin{equation}
  \begin{split}
  \bm \tau = & G\frac{\mu_\mathcal{A}\mu_\mathcal{B}}{2}\left[\sum_{\ell m} a_\mathcal{B}^\ell J_{\ell m} \sum_{\ell' m'}a_\mathcal{A}^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D) (-1)^{\ell'}\right.\\
  \times & \left.\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(\alpha, \beta, \gamma)^* \right. \\
  \times & \Big((i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} \\
  & +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\Big) \Bigg],
  \end{split}
  \label{eqn:tidal-torque}
\end{equation}
where $\bm D$ is the position of the asteroid; $\alpha$, $\beta$, and $\gamma$ are Euler angles expressing the orientation of the asteroid; and  $\mu_\mathcal{B}$ and $J_{\ell m}$ are the mass and density moments of the central body respectively; and $S_{\ell m}$ are the irregular solid spherical harmonics.

In section \ref{sec:sim}, we describe the simulation used to integrate the equations of motion and ins section

@@

\include{appendices}
